---
title: "Tidy lipidomics data"
output:
  html_document:
    df_print: paged
params:
  input_folder: AMV
  output_folder: ./
  metadata_name: metadata
---

## Load packages

```{r tidyverse and janitor, message=FALSE}
library(tidyverse)
library(janitor)
```

## Import data

```{r Import meta and LV data, message=FALSE, warning=FALSE}
# Paths
input_folder <- params$input_folder
output_folder <- params$output_folder
metadata_name <- params$metadata_name

# Import meta data
meta_data <- read.csv(list.files(path = input_folder, pattern = metadata_name, full.names = T)) %>% 
  clean_names()

# Import LipidView data
files <- list.files(path = input_folder, pattern = ".txt", full.names = TRUE) 

# Parse lipid class from file names
splitted_files <- str_split(files, pattern = "_", simplify = TRUE)
names(files) <- str_replace_all(splitted_files[,ncol(splitted_files)],".txt", "")

raw_data <- map_df(files, read_tsv, skip = 1, .id = "class") %>% 
  clean_names()
```

## Tidy LipidView data

```{r clean and prepare metadata}
# Extract sample input [ul] from meta data
sample_input <- meta_data %>%
  filter(grepl("sample input", class)) %>% 
  select(-class, -rf_values, -x) %>%
  pivot_longer(everything(),
               names_to = "samples",
               values_to = "sample_input")


# Extract internal standard concentrations [pmol]
# and response factors (rf-values) from meta data
standard_input <- meta_data %>%
  filter(!grepl("sample input", class)) %>% 
  select(-x) %>%
  pivot_longer(
    cols = c(-class, -rf_values),
    names_to = "samples",
    values_to = "standard_input")

```

```{r clean and prepare LipidView files}
# Tidy up LipidView-files and join LipidView data with meta data
raw_data <- raw_data %>%
  filter(!grepl("Sample ID", sample_name)) %>% 
  select(-sample_name, -x2, -x3, -x4) %>% 
  rename("species" = x5,
         "scan_name"= x6) %>% 
  pivot_longer(
    cols = c(-species, -class, -scan_name),
    names_to = "samples",
    values_to = "intensity") %>% 
  mutate(species = str_remove(species, "\\+NH4"))


# Seperate internal standards with samples and return the mean of IS-intensities to the rawdata
is_data <- raw_data %>% 
  filter(str_detect(species, "IS"),
         class != "TAG") %>% 
  group_by(class, samples) %>% 
  summarise(is_intensity = mean(intensity))

# Special case for TAG
is_tag_data <- raw_data %>% 
  filter(str_detect(species, "IS"),
         species == "IS D5-TAG 17:0/17:1/17:0",
         str_detect(scan_name, "17:0")) %>% 
  select(class, samples, is_intensity = intensity)

is_data <- full_join(is_data, is_tag_data, by = c("class", "samples", "is_intensity"))
```

```{r bring all data together}
# 
tidy_data <- raw_data %>% 
  select(-scan_name) %>% 
  filter(!str_detect(species, "IS")) %>% 
  left_join(is_data, by = c("class", "samples")) %>% 
  left_join(standard_input, by = c("class", "samples")) %>% 
  left_join(sample_input, by = "samples")
```

### Special cases

Before the script can proceed with number crunching, there are some special cases to wdeal with. These cases are classes with NA in rf_values (e.g. Cholesterol and Ergosterol), DAG and TAG.
 
- Cholesterol & Ergosterol: Calculate the rf-value with kon_b and kon_c
- DAG: Sum DAG-species with the same fatty acid sum, but different fatty acid composition. 
       Better way is to calculate the possible sum of detected fatty acids in neutral loss-scans.
- TAG: Same as DAG, but with 3 Fatty Acids

But for now, simple sums of the same DAG- and TAG-species (FA_sum) are sufficient.

```{r Calc missing rf-values}
tidy_data %>%
  filter(grepl("kon", samples) & is.na(rf_values)) %>% 
  mutate(pmol = standard_input/is_intensity * intensity)
  

```














