---
title: "LipidView Analysis"
output:
  html_document:
    df_print: paged
params:
  input_folder: AMV
  output_folder: ./
  metadata_name: metadata
editor_options: 
  chunk_output_type: console
---

# Load packages and functions

```{r Tidyverse and janitor, message=FALSE}
library(tidyverse)
library(janitor)
library(openxlsx)
library(ggpubr)

source("LipidView_analysis_funs.R")
```

# Import data

```{r Import meta and LV data, message=FALSE, warning=FALSE}
# Paths
input_folder <- params$input_folder
output_folder <- params$output_folder
metadata_name <- params$metadata_name

# Import meta data
meta_data <- read.csv(list.files(path = input_folder,
                                 pattern = metadata_name,
                                 full.names = T)) %>%
  clean_names()

# List all (LipidView) txt.files in designated input_folder
files <- list.files(path = input_folder, pattern = ".txt", full.names = TRUE)

# Parse lipid class from file names
splitted_files <- str_split(files, pattern = "_", simplify = TRUE)
names(files) <- str_remove_all(splitted_files[, ncol(splitted_files)], ".txt")

# Read and merge all LipidView txt.files
raw_df <- map_df(files, read_tsv, skip = 1, .id = "class") %>%
  clean_names()
```

# Prepare data for arithmetic operations

## Prepare LipidView data

```{r Clean and prepare LipidView files}
# General clean up of the LipidView dataframe:
## Get rid off unnecessary columns and strings, rename columns
raw_data <- raw_df %>%
  filter(!str_detect(sample_name, "Sample ID")) %>%
  select(-sample_name, -x2, -x3, -x4, -x6) %>%
  rename("species" = x5) %>%
  pivot_longer(
    cols = c(-species, -class),
    names_to = "samples",
    values_to = "intensity") %>%
  mutate(species = str_remove_all(species, "\\+NH4"))

# Extract the intensities of internal standards and
# return the mean of IS-intensities to the rawdata
is_data <- raw_data %>%
  filter(str_detect(species, "IS"),
         class != "TAG") %>%
  group_by(class, samples) %>%
  summarise(is_intensity = mean(intensity))

# Special case for TAG:
# Because "IS D5-TAG 17:0/17:1/17:0" is the only ISTD for
# TAG quantification in use, it is filtered by hardcode for now.
is_data <- raw_data %>%
  filter(str_detect(species, "IS"),
         species == "IS D5-TAG 17:0/17:1/17:0") %>%
  group_by(class, samples) %>%
  summarise(is_intensity = sum(intensity)) %>%
  full_join(is_data, by = c("class", "samples", "is_intensity"))
```

## Prepare meta data

```{r Clean and prepare metadata}
# Extract sample input [ul] from meta data
sample_input <- meta_data %>%
  filter(class == "sample input") %>%
  select(-class, -rf_values, -x) %>%
  pivot_longer(everything(),
               names_to = "samples",
               values_to = "sample_input")

# Extract internal standard concentrations [pmol]
# and response factors (rf-values) from meta data
standard_input <- meta_data %>%
  filter(class != "sample input") %>%
  select(-x) %>%
  pivot_longer(
    cols = c(-class, -rf_values),
    names_to = "samples",
    values_to = "standard_input")
```

## Join them all

```{r Bring all data together, message=FALSE, warning=FALSE}
# Join data and get rid off redundant cols (scan_name)
# and observations (IS rows)
joined_data <- raw_data %>%
  left_join(is_data, by = c("class", "samples")) %>%
  left_join(standard_input, by = c("class", "samples")) %>%
  left_join(sample_input, by = "samples") %>%
  filter(!str_detect(species, "IS"))

#To preserve order of species after summarise-function
joined_data$species <- fct_inorder(joined_data$species, ordered = NA)
```

## Special cases

Before the script can proceed with number crunching, there are some special cases to deal with. Those are classes without predetermined rf_values (e.g. Cholesterol & Ergosterol) and classes with identical species from different scans(e.g. neutral loss-scans of fatty acids in DAG & TAG) 

Tasks:

- classes without predetermined rf_values: Calculate the rf-value with kon_b and kon_c

- classes with identical species from different scans: Sum DAG/TAG-species with the same fatty acid sum, but different fatty acid composition. 
  - Better way is to calculate the possible sum of detected fatty acids in neutral loss-scans.
  - But for now, simple sums of the same DAG- and TAG-species (FA_sum) are sufficient.

#### Classes without predetermined rf_values

```{r Calc missing rf-values, message=FALSE, warning=FALSE}
if (any(is.na(joined_data$rf_values)) == TRUE) {

  # Calculate pmol-values of kon_a, _b and _c
  missing_rf <- joined_data %>%
    filter(str_detect(samples, "kon") & is.na(rf_values)) %>%
    mutate(pmol = (standard_input / is_intensity * intensity))

  # create kon_a variable for better reading
  kon_a <- missing_rf %>%
    filter(samples == "kon_a") %>%
    select(class, species, kon_a_pmol = pmol)

  # add new column with pmol-values of kon_a for blank substraction and
  missing_rf <- left_join(missing_rf, kon_a, by = c("class", "species")) %>%
    mutate(rf_values = sample_input / (pmol - kon_a_pmol)) %>%
    filter(!str_detect(samples, "kon_a")) %>%
    group_by(class, species) %>%
    summarise(rf_values = mean(rf_values))

  # join and unite the calculated rf-values and get rid off kon_b and c.
  # unite()-function merges 2 columns and returns characters.
  # Therefore remove NA-strings and convert back to numeric)
  joined_data <- left_join(joined_data,
                           missing_rf,
                           by = c("class", "species")) %>%
    unite("rf_values",
          "rf_values.y",
          "rf_values.x",
          sep = "",
          remove = TRUE,
          na.rm = TRUE) %>%
    mutate(rf_values = str_remove_all(rf_values, "NA"),
           rf_values = as.numeric(rf_values)) %>%
    filter(!(samples %in% c("kon_b", "kon_c")))

}
```

#### Classes with identical species from different scans

```{r Sum identical species}
data <- joined_data %>%
  group_by_at(vars(-intensity)) %>%
  summarise(intensity = sum(intensity)) %>%
  ungroup()
```

# Number crunching

## Apply the Response factor (rf-values) to endogenous intensities

```{r Apply the rf-value}
data <- data %>%
  mutate(intensity = intensity * rf_values) %>%
  select(-rf_values)
```

## Quantify lipid species

#### Convert intensities to picomoles

```{r [intensities] to [pmol]}
data <- data %>%
  mutate(pmol = (standard_input / is_intensity) * intensity) %>%
  select(-is_intensity, -intensity)
```

#### Blank (kon_a) substraction

```{r sample[pmol] - kon_a[pmol]}
# create kon_a variable for better reading
kon_a <- data %>%
  filter(samples == "kon_a") %>%
  select(class, species, kon_a_pmol = pmol)

data <- left_join(data, kon_a, by = c("class", "species")) %>%
  mutate(pmol_corr = pmol - kon_a_pmol,
         pmol_corr = ifelse(pmol_corr < 0, 0, pmol_corr)) %>%
  filter(!str_detect(samples, "kon_a")) %>%
  select(-kon_a_pmol)
```

#### Convert picomoles to micromolar

```{r [pmol] to [uM]}
data <- data %>%
  mutate(uM = pmol_corr / sample_input)
```

# Create the Excel workbook

## Prepare Excel workbook

```{r Create Excel wb and set font}
wb <- createWorkbook()
modifyBaseFont(wb, fontSize = 10, fontColour = "black", fontName = "Arial")
```

## Write meta_data worksheet

```{r Create meta_data worksheet}
addWorksheet(wb, "metadata")

setColWidths(wb, "metadata", col = 1:99, widths = 10)
setRowHeights(wb, "metadata", rows = c(1:99), heights = 15)
  
writeData(wb, "metadata", meta_data, startCol = 1, startRow = 1)
```

## Write lipid class-worksheets

```{r Create one worksheet for each lipid class}
lipid_classes <- unique(data$class)

for (f in lipid_classes) {

  addWorksheet(wb, f)

  setColWidths(wb, f, col = 1, widths = 20)
  setColWidths(wb, f, col = 2:99, widths = 10)
  setRowHeights(wb, f, rows = 2, heights = 50)
  setRowHeights(wb, f, rows = c(1, 3:99), heights = 15)

  # main-table with sample/meta data and summed micromolar/picomole values
  excel_sum_data <- data %>%
    filter(class == f) %>%
    group_by(samples, sample_input, standard_input) %>%
    summarise(sum(pmol),
              sum(pmol_corr),
              sum(uM))

  colnames(excel_sum_data) <- c("sample number",
    paste("sample input [ul]"),
    paste(f, "IS input [pmol]"),
    paste(f, "[pmol] before corr"),
    paste(f, "[pmol] after corr"),
    paste(f, "[uM]"))

  # summed micromolar-table splitted by diacyl- and ether/acyl-species
  ether <- data %>%
  filter(class == f) %>%
  mutate(ether = str_detect(species, "O-")) %>%
  group_by(samples, ether) %>%
  summarise(sum(uM)) %>%
  pivot_wider(names_from = ether,
              values_from = "sum(uM)")

  colnames(ether) <- c("sample number",
                       paste0("a", f, " [uM]"),
                       if ("TRUE" %in% colnames(ether)) {
                         paste0("e", f, " [uM]")
                       }
                      )

  # mol%-table (lipid species profile)
  excel_species_profile <- data %>%
    filter(class == f) %>%
    group_by(class, samples) %>%
    mutate(uM_sum = sum(uM)) %>%
    mutate(mol_percent = uM * 100 / sum(uM),
           species = str_remove_all(species, f)) %>%
    ungroup %>%
    select(samples, species, mol_percent) %>%
    pivot_wider(names_from = species,
                values_from = mol_percent) %>%
    rename(" " = samples)

  # all-in-one plot
  print(
    ggarrange(lipid_barplot(data, f, "species"),
      ggarrange(lipid_barplot(data, f, "chain_length"),
                lipid_barplot(data, f, "dbonds"),
                lipid_barplot(data, f, "oh"),
                ncol = 3, common.legend = TRUE, legend = "bottom"),
    nrow = 2, legend = "none")
  )

  # write previous created df to every sheet via function for better reading
  # at the end of this fun the printed plot will be inserted to the sheet
  write_data_to_excelsheet(excel_sum_data,
                           ether,
                           excel_species_profile)
}

```

## Write summary-worksheet for class profile

```{r Create summary-sheet}
class_order <- c("samples",
                 "sample_input",
                 "PC", "ePC",
                 "LPC",
                 "SM",
                 "PE", "ePE",
                 "PS", "ePS",
                 "PI", "ePI",
                 "PG", "ePG",
                 "PA", "ePA",
                 "plPE16-0", "PE P-16-0",
                 "plPE18-1", "PE P-18-1",
                 "plPE18-0", "PE P-18-0",
                 "Cer",
                 "tCer",
                 "HexCer", "GlcCer",
                 "Hex2Cer", "LacCer",
                 "EE",
                 "SAA", "Ergo",
                 "CA", "Chol",
                 "CE",
                 "DAG",
                 "TAG",
                 "sum [uM]",
                 "sum [pmol]")

excel_class_profile <- data %>%
  mutate(ether = str_detect(species, "O-")) %>%
  group_by(class, samples, ether, sample_input) %>%
  summarise("[uM]" = sum(uM), .groups = "drop") %>%
  mutate(class = paste0(ether, "_", class),
         class = str_remove(class, "FALSE_"),
         class = str_replace(class, "TRUE_", "e")) %>%
  select(-ether) %>%
  group_by(samples) %>%
  mutate("sum [uM]" = sum(`[uM]`),
         "sum [pmol]" = `sum [uM]` * sample_input) %>%
  pivot_wider(names_from = "class",
              values_from = "[uM]") %>%
  select(class_order[class_order %in% colnames(.)])


addWorksheet(wb, "summary")

setColWidths(wb, "summary", col = 1, widths = 20)
setColWidths(wb, "summary", col = 2:99, widths = 10)
setRowHeights(wb, "summary", rows = 2, heights = 50)
setRowHeights(wb, "summary", rows = c(1, 3:99), heights = 15)

writeData(wb, "summary", excel_class_profile, 
          startCol = 1, startRow = 2, rowNames = FALSE)
```

## Build the Excel-file

```{r Define Sheet-order and build excel workbook}
excel_sheet_order <- c("metadata",
                       class_order[class_order %in% sheets(wb)],
                       "summary")

worksheetOrder(wb) <- match(excel_sheet_order, sheets(wb))

saveWorkbook(wb, "test.xlsx", overwrite = TRUE)
```